---
title: "Automation with makefiles"
---

## Introduction

This section is aiming for a happy medium of flexibility for producing training material. The approach is inspired by how Software Carpentry structure their git repositories, however by using the `rmarkdown` package we can keep things simple. If you want to go simpler still, `rmarkdown` also has a [`render_site`](http://rmarkdown.rstudio.com/rmarkdown_websites.html) function which can build a whole website at once, which might be sufficient to your needs.

We will first look at the `rmarkdown` package, then how to use this package from a Makefile to automate building of .html files from .Rmd files.

## rmarkdown package

Under the hood, when you press the "Knit" button in RStudio the `rmarkdown` package's `render` function is used. We can also use this package from the R console.

```{r eval=FALSE}
rmarkdown::render("foo.Rmd")
```

The `rmarkdown` package in turn uses the `knitr` package to actually run R code, and external program `pandoc` to create HTML. When creating a PDF, `pandoc` in turn uses `pdflatex`.


### rmarkdown from the shell

We can run this from a shell using:

```{r eval=FALSE, highlight=FALSE}
Rscript -e 'rmarkdown::render("foo.Rmd")'
```


## Makefiles

A programmer's natural instinct will be to place this in a "Makefile" so that the `make` command can make the HTML document whenever we need to.

### Rules

Create a file called `Makefile` (in RStudio, select "File/New File/Text File", and save it with the name `Makefile`). Enter the following in the file:

```{r eval=FALSE, highlight=FALSE}
foo.html : foo.Rmd
    Rscript -e 'rmarkdown::render("foo.Rmd")'
```

**Important:** The indentation in a Makefile is important, and it must be a tab character. You can check this from R with the command `readLines("Makefile")` -- you should see the indent shown as a `\t` escape sequence. 

We have described how to create the file `foo.html` from the input file `foo.Rmd`. Multiple input files can be given if necessary, separated by spaces.

Now in a shell, type:

```{r eval=FALSE, highlight=FALSE}
make foo.html
```

Nothing happens, as the `.html` file was created after out last edit of the `.Rmd`. Make thinks the file is up to date, and does not run the command.

Edit the `.Rmd` file, and again run `make foo.html` in the shell. This time the command should run.

If you just type `make` on its own, the first rule in the `Makefile` is used.


#### Exercise {.challenge}

Add some code to `foo.Rmd` that deliberately contains an error. What happens when you try to `make` it?


### "Pattern" rules

Create a second R Markdown file, `bar.Rmd`. We could add a second rule to the Makefile to convert this into HTML, but this becomes repetitive. A better option is the use a "pattern rule":

```{r eval=FALSE, highlight=FALSE}
%.html : %.Rmd
    Rscript -e 'rmarkdown::render("$<")'
```

Note here:

* `%` symbol in the input and output filenames indicates this is a pattern rule.
* In the command we've use the special variable `$<` to insert the input filename into the command. More on variables later.

Now to render our second file we can type:

```{r eval=FALSE, highlight=FALSE}
make bar.html
```

### A "phony" rule to make all files

Consider this Makefile:

```{r eval=FALSE, highlight=FALSE}
all : foo.html bar.html
    echo All files are now up to date

%.html : %.Rmd
    Rscript -e 'rmarkdown::render("$<")'
```

* The `all` rule depends on some files, so `make` will ensure these are up to date. 

* The command `echo` in the rule just prints out a message (this can even be left out entirely). `make` does not check that a file called `all` was actually created.

* The `all` rule is the first rule in the `Makefile`, so it is the default rule that is run when you type `make`.

Try editing one of your Rmarkdown files and running `make` again. Notice only the command to build that file's corresponding HTML file is run.


### Cleaning up

It's common practice to also have a rule to clean away all the files that have been created. This gives a way to test everything works from a blank slate.

```{r eval=FALSE, highlight=FALSE}
all : foo.html bar.html
    echo All files are now up to date

clean : 
    rm -f foo.html bar.html 

%.html : %.Rmd
    Rscript -e 'rmarkdown::render("$<")'
```

Now we can type `make clean` to clean up, and `make clean ; make` will test that everything can be built without errors.


### Variables

A final element of a Makefile is use of variables. With this, you should be able to understand most of what is going on in Makefiles you encounter in the wild. 

In our last Makefile, we typed `foo.html bar.html` twice. It would be nice to avoid this repetition. This can be achieved with a variable:

```{r eval=FALSE, highlight=FALSE}
HTML_FILES=foo.html bar.html

all : $(HTML_FILES)
    echo All files are now up to date

clean : 
    rm -f $(HTML_FILES) 

%.html : %.Rmd
    Rscript -e 'rmarkdown::render("$<")'
```


#### Exercise {.challenge}

Create a file `index.Rmd`, with hyperlinks to `foo.html` and `bar.html`. Recall that hyperlinks can be created in markdown with `[link text](destination.html)`.

Adjust your Makefile so this file is also built into an HTML file.

Now you have a website. To test your website, from the "Files" pane in RStudio, click on `index.html` and then "View in Web Browser".

You could make this public for example by using [GitHub Pages](https://pages.github.com/), or by running your own webserver (such as [Apache](https://httpd.apache.org/) or [nginx](https://nginx.org/en/)).


## Python

The discussion above has focussed on R Markdown. In Python, a similar role is played by [Jupyter](http://jupyter.org/) notebooks. These are created in a web-based editor, which can be started by typing `jupyter notebook` in the shell. It is also possible to run a multi-user [Jupyter Hub](https://jupyterhub.readthedocs.io/en/latest/) on a server. Notebooks are stored in JSON format, and have extension `.ipynb`.

Suppose you have created a notebook called `my_notebook.ipynb`. Then the following shell command can be used to run the code in it and convert the notebook to HTML:

```{r eval=FALSE, highlight=FALSE}
jupyter nbconvert --execute --to html my_notebook.ipynb
```

One difference between `.Rmd` and `.ipynb` files is that `.ipynb` files also contain cached output of code chunks. Specifying the `--execute` option here ensures all of the code in the notebook is run afresh. Using this command in a Makefile would ensure the build will halt with an error if there are any problems with the code.



## Discussion

`make clean ; make` can be used to test all of the code in our tutorial. We will see an error message and the process will abort immediately if anything goes wrong as the HTML files are built. This is also a handy way to test that a training server has all the necessary R or Python packages installed!

You might want to add some further rules, for example:

* Creating a ZIP file of files used in the tutorial for easy download.
* Creating a PDF verions of the notes, by creating markdown files using either `rmarkdown` or `knitr` and then using `pandoc` to create a PDF using LaTeX.
* Running scripts to save results of lengthy calculations, so that rendering Rmd files remains fast.

If someone wants to adapt our tutorial, they can easily do so. Having made the changes they want, the HTML files can be re-built with a single command. Give your notes an appropriate licence, such as a [Creative Commons](https://creativecommons.org/) license, so that people know they can do this without having to ask your permission.

Setting up tutorial notes in this way gives us desirable features of open science. We have left nothing out or the code will not run. They are easily reproducable and sharable. They are written plain text formats that anyone can edit, and built with tools that are all freely available. They can serve as a base that anyone who wants to can build on.







